Abstraction and Invariance for Algebraically Indexed Types

### 2012-10-22 paper/section-metric.tex
Update the distance-indexed types section to new rel. envs.

\autoref{sec:instantiations} and \autoref{sec:monoid-indexed-types}
presented examples where the relational interpretations of primitive
types relate pairs of elements by some transformation if applying the
transformation to the first element of a pair yields the second. Thus,
the free theorems that we derived directly take the form of
``invariance'' properties, where some equation holds between two
terms. In this section, we examine another instantiation of our
general framework that relates values when they are within a certain
distance. The free theorems that we obtain inform us of the effect
that programs have on the distances between values. For example, a
program $M$ of type:
\begin{displaymath}
  \forall \epsilon_1,\epsilon_2\mathord:\metricSort.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2}
\end{displaymath}
must satisfy the property that for all $\epsilon_1, \epsilon_2 > 0$
and $x, x', y, y' \in \mathbb{R}$:
\begin{displaymath}
  \begin{array}{l}
    \textrm{if }|x - x'| < \epsilon_1 \textrm{ and } |y - y'| < \epsilon_2 \textrm{ then}\\
    \hspace{2em}|\tmSem{M}\ \eta_\Metric\ x\ y - \tmSem{M}\ \eta_\Metric\ x'\ y'| < \epsilon_1 + \epsilon_2
  \end{array}
\end{displaymath}

@@

For the families of sets of relational environments
$\relEnv{E}_\Metric$, we use the model-based construction in
\autoref{sec:constr-rel-env}, interpreting the primitive type
$\tyPrimNm{real}$ with $R_{\tyPrimNm{real}}(\epsilon) = \{ (x,x')
\sepbar |x - x'| < \epsilon\}$.
%\begin{displaymath} %FIXME: define \Eq earlier on
%  \begin{array}{l@{\hspace{2em}}l}
%    R^\bullet_{\tyPrimNm{real}} = \Eq_\mathbb{R} &
%    R_{\tyPrimNm{real}}(\epsilon) = \{ (x,x') \sepbar |x - x'| < \epsilon \}
%  \end{array}
%\end{displaymath}
\begin{lemma}\label{lem:metric-environments}
  For all $\Delta$ and $\rho \in \relEnv{E}_\Metric(\Delta)$,
  $(\eta_\Metric, \eta_\Metric) \in
  \rsem{\Gamma}{\relEnv{E}_\Metric}{\rho}$.
\end{lemma}

\paragraph{Uniform continuity}
  %By making use of 
Using %the
existential types, % in our framework,
%we can state 
the standard $\epsilon$-$\delta$ definition of
\emph{uniform} continuity can be expressed as %a type:
%\begin{displaymath}
$  \forall \epsilon \mathord: \metricSort.\ \exists \delta\mathord: \metricSort.\\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}$.
%\end{displaymath}
For any program $M$ of this type, %we obtain from
\thmref{thm:abstraction} gives a free theorem that is exactly 
%the definition of 
uniform continuity:
\begin{displaymath}
  \forall \epsilon \mathord> 0. \exists \delta \mathord> 0. \forall x,x'. |x - x'| < \delta \Rightarrow |\tmSem{M}\ x - \tmSem{M}\ x'| < \epsilon
\end{displaymath}
This definition differs from the $\epsilon$-$\delta$ definition of
(regular) continuity by the order of quantification:
  %For continuity, 
there, $\forall x$ comes before $\exists \delta$, so the distance
$\delta$ may depend on %the value of
$x$. We suspect that to express standard continuity as a type would
require some form of type dependency. Chaudhuri, Gulwani and
Lublinerman \cite{chaudhuri10continuity} have given a program logic
based approach to verifying the continuity of programs.

>>>>

\autoref{sec:instantiations} and \autoref{sec:monoid-indexed-types}
presented examples where the relational interpretations of primitive
types relate pairs of elements by some transformation if applying the
transformation to the first element of a pair yields the second. Thus,
the free theorems that we derived directly take the form of
``invariance'' properties, where some equation holds between two
terms. In this section, we examine another instantiation of our
general framework that relates values when they are within a certain
distance. The free theorems that we obtain inform us of the effect
that programs have on the distances between values. For example, a
program $M$ of type:
\begin{displaymath}
  \forall \epsilon_1,\epsilon_2\mathord:\metricSort.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2}
\end{displaymath}
must satisfy the property that for all $\epsilon_1, \epsilon_2 > 0$
and $x, x', y, y' \in \mathbb{R}$:
\begin{displaymath}
  \begin{array}{l}
    \textrm{if }\abs{x - x'} < \epsilon_1 \textrm{ and } \abs{y - y'} < \epsilon_2 \textrm{ then}\\
    \hspace{2em}\abs{\tmSem{M}\ \eta_\Metric\ x\ y - \tmSem{M}\ \eta_\Metric\ x'\ y'} < \epsilon_1 + \epsilon_2
  \end{array}
\end{displaymath}

@@

For the relational intepretation we construct a model of the indexing
theory by interpreting $\metricSort$ with strictly positive real
numbers. We set $\rsem{\tyPrimNm{real}}(\epsilon) = \{ (x,x') \sepbar
\abs{x - x'} < \epsilon \}$.
\begin{lemma}\label{lem:metric-environments}
  For all $\Delta$ and $\rho \in \rsem{\Delta}$, $(\eta_\Metric,
  \eta_\Metric) \in \rsem{\Gamma}{\rho}$.
\end{lemma}

\paragraph{Uniform continuity}
  %By making use of 
Using %the
existential types, % in our framework,
%we can state 
the standard $\epsilon$-$\delta$ definition of
\emph{uniform} continuity can be expressed as %a type:
%\begin{displaymath}
$  \forall \epsilon \mathord: \metricSort.\ \exists \delta\mathord: \metricSort.\\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}$.
%\end{displaymath}
For any program $M$ of this type, %we obtain from
\thmref{thm:abstraction} gives a free theorem that is exactly 
%the definition of 
uniform continuity:
\begin{displaymath}
  \forall \epsilon \mathord> 0. \exists \delta \mathord> 0. \forall x,x'. abs{x - x'} < \delta \Rightarrow \abs{\tmSem{M}\ x - \tmSem{M}\ x'} < \epsilon
\end{displaymath}
This definition differs from the $\epsilon$-$\delta$ definition of
(regular) continuity by the order of quantification:
  %For continuity, 
there, $\forall x$ comes before $\exists \delta$, so the distance
$\delta$ may depend on %the value of
$x$. We suspect that to express standard continuity as a type would
require some form of type dependency. Chaudhuri, Gulwani and
Lublinerman \cite{chaudhuri10continuity} have given a program logic
based approach to verifying the continuity of programs.
